Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2017-08-04 manuscript/manuscript.tex
minor edits

The primary data analysis technique used by PTAs is Bayesian inference 
to compare the probabilities of various models for the pulsar timing residuals, 
including one where the residuals contain evidence of an isotropic GW stochastic background. 
Bayesian inference is a powerful technique because it allows us to compare arbitrarily complicated 
models; however, they are computationally intensive, 
particularly when searching for evidence of dipole spatial correlations -- 
``the smoking gun'' of a stochastic GWB.

The significance of a stochastic GWB can also be assessed using the 
optimal statistic, a frequentist estimator for the GWB amplitude. 
It is an important complement to Bayesian detection techniques 
because it provides an independent detection procedure, 
and it is significantly faster to compute. 
In particular, different spatial correlations can be tested with the optimal statistic 
in less than a second, 
whereas a Bayesian analysis with spatial correlations requires 
several weeks to run on a supercomputing cluster.

@@

We used these three techniques to compute the optimal statistic for simulated data sets. 
We constructed simulated data sets using the same procedure as in \citet{tlb+2017}. 
The simulated PTAs consisted of 10 MSPs with 
total observation times and noise properties chosen to be 
similar to the \sv{Steve fill in here...} (see Table~\ref{tab:sim}). 
We also injected a stochastic GWB amplitude with $\Agw = 5\times10^{-15}$.
\begin{table}[htb]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTA}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $\log_{10}A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0613$-$0200 & 16.054 & 1.58 & -13.90 & 3.18 \\
		J0751+1807 & 17.606 & 2.60 & -14.14 & 2.58 \\
		J1012+5307 & 16.831 & 1.47 & -13.09 & 1.65 \\
		J1640+2224 & 16.735 & 1.99 & -13.24 & 0.03 \\
		J1643$-$1224 & 17.300 & 1.65 & -18.56 & 4.04 \\
		J1713+0747 & 17.657 & 0.26 & -14.90 & 4.85 \\
		J1744$-$1134 & 17.250 & 0.65 & -13.60 & 2.00 \\
		J1857+0943 & 17.310 & 1.51 & -16.00 & 1.35 \\
		J1909$-$3744 & 9.379 & 0.12 & -13.99 & 2.06 \\
		J2145$-$0750 & 17.161 & 1.19 & -13.87 & 4.02 \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}

>>>>

The primary data analysis technique used by PTAs is Bayesian inference 
to compare the probabilities of various models for the pulsar timing residuals, 
including one where the residuals contain evidence of an isotropic GW stochastic background \sv{add citation}. 
Bayesian inference is a powerful technique because it allows us to compare arbitrarily complicated models. 
However, running a full Bayesian analysis is computationally intensive, 
particularly when searching for evidence of dipole spatial correlations -- 
``the smoking gun'' of a stochastic GWB.

The significance of a stochastic GWB can also be assessed using the 
optimal statistic, a frequentist estimator for the GWB amplitude \sv{add citation}. 
It is an important complement to Bayesian detection techniques 
because it provides an independent detection procedure, 
and it is significantly faster to compute. 
In particular, different spatial correlations can be tested with the optimal statistic 
in less than a second, 
whereas a Bayesian analysis with spatial correlations requires 
several weeks to run on a supercomputing cluster.

@@

We used these three techniques to compute the optimal statistic for simulated data sets. 
We constructed simulated data sets using the same procedure as in \citet{tlb+2017}. 
The simulated PTAs consisted of 10 MSPs with 
total observation times and noise properties chosen to be 
similar to the properties of the EPTA (see Table~\ref{tab:sim})  \sv{add citation}. 
We also injected a stochastic GWB amplitude with $\Agw = 5\times10^{-15}$.
\begin{table}[htb]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTA}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $\log_{10}A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0613$-$0200 & 16.054 & 1.58 & -13.90 & 3.18 \\
		J0751+1807 & 17.606 & 2.60 & -14.14 & 2.58 \\
		J1012+5307 & 16.831 & 1.47 & -13.09 & 1.65 \\
		J1640+2224 & 16.735 & 1.99 & -13.24 & 0.03 \\
		J1643$-$1224 & 17.300 & 1.65 & -18.56 & 4.04 \\
		J1713+0747 & 17.657 & 0.26 & -14.90 & 4.85 \\
		J1744$-$1134 & 17.250 & 0.65 & -13.60 & 2.00 \\
		J1857+0943 & 17.310 & 1.51 & -16.00 & 1.35 \\
		J1909$-$3744 & 9.379 & 0.12 & -13.99 & 2.06 \\
		J2145$-$0750 & 17.161 & 1.19 & -13.87 & 4.02 \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}
